---
title: "Assignment 2"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Exercise 1  
Read the data in the chunk below. Propose a statistical model (using mathematical notation) to describe the relationship between corn yield and plant density, and fit that model to the data.  

$$y = ...$$

```{r}
library(tidyverse)
url <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/corn_example2.csv"
data <- read.csv(url)
data %>% 
  ggplot(aes(plant_density, yield_Mgha))+
  geom_point()+
  labs(x = expression(Plant~Density~(plants~m^{-2})), 
       y = expression(Grain~Yield~(Mg~ha^{-1})))+
  theme_classic()+
  coord_cartesian(xlim = c(0, 15), 
                  ylim = c(0, 16), 
                  expand = F)+
  theme(aspect.ratio = 1)
```

### Answer the following questions:  

a. What is the plant density that maximizes grain yield? Provide a point estimate and some measure of uncertainty.  
b. How much yield do you expect the crop to yield **on average** with 8 plants per $\text{m}^2$? What is a good 95% confidence interval for that value?    

c. What is a reasonable 95% confidence interval for observable yields at 8 plants per $\text{m}^2$?    

d. Which confidence interval would be affected most is the sample size was increased twofold?    

