---
title: "Homework 4"
author: "J Lacasa"
date: "2024-10-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Read the data in the chunk below. The data can be also found [here](https://www.kaggle.com/datasets/soujanyahp/prostate-cancer-dataset).  

The data come from [a study](https://www.sciencedirect.com/science/article/abs/pii/S002253471741175X) that aims to "*examine the relationship of preoperative ambulatory serum prostate specific antigen (PSA) to cancer volume, capsular penetration into the periprostatic fat, seminal vesicle invasion, margin-positive tissue planes, lymph node metastases, Gleason grade of the prostate cancer, prostate weight and amount of benign prostatic hyperplasia (BPH) within the radical prostatectomy specimen*".

Please provide all your code in order to make your results reproducible and submit an html od pdf file to canvas by October 31st.  

```{r}
url <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/prostate.csv"
dd_prostate <- read.csv(url)
names(dd_prostate)
```

### 1. Statistical model  
Design a statistical model that describes log PSA (`lpsa` column in the dataframe) as a function of at least 3 of the predictor variables, and provide the estimates (i.e., the values for your $\hat{\beta}$). Fit the statistical model using only the data that contain `train = TRUE`, and evaluate the model with the data that contain `train = FALSE`. 

Please mention:  
1. The variable selection criteria you use.  
2. The estimates (i.e., the values for your $\hat{\beta}$) you obtained for your final model.  

*Note: there are many ways to complete this task. As long as it is correctly justified, you can pick any option we studied in class.*

### 2. Results  

Interpret and describe your results in 3 sentences. For example: mention the relationship between the different predictors and the response lpsa (e.g., positive or negative), mention the relative importance of the different predictors to explain changes in lpsa (e.g., variable $x_1$ showed a stronger relationship to lpsa than $x_2$).    



