---
title: "Day 7 - 09/04/2024"
format: html
editor: visual
---

## From last class

-   What we know so far: estimation\
-   Cool properties of $\hat{\beta}$: unbiased, minimum variance.

## Uncertainty estimation

-   The importance of reporting uncertainty estimates.\
-   A funny example about reporting point estimates without uncertainty estimates.\
-   Who has heard about meta-analyses? \[[book example](https://link.springer.com/book/10.1007/978-94-024-1696-1)\]

### Estimation uncertainty versus prediction uncertainty

How is the data generated? For the model

$$y_i \sim N(\mathbf{x}_i' \boldsymbol{\beta}, \sigma^2),$$

we have two sources of uncertainty:

-   From estimating ${\boldsymbol{\beta}}$: $\hat{\boldsymbol{\beta}} \sim N(\boldsymbol{\beta}, \frac{\sigma^2}{(n-1)s^2_x})$ ([estimation uncertainty])
-   From observing new data: $y_{new} \sim N(\mathbf{x}_{new}' \boldsymbol{\beta}, \sigma^2)$ ([prediction uncertainty])

#### Estimation uncertainty

Recall the clover example

```{r message=FALSE, warning=FALSE}
library(tidyverse)
url <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/lotus_hw1.csv"
data <- read.csv(url)
m <- lm(stm.length_cm ~ 1 + doy,  data = data)
confint(m)
```

The 95% confidence interval for $\beta_1$, (`r round(confint(m)[2], 2)`, `r round(confint(m)[4], 2)`), means that we can say that **the slope** of length vs. doy (i.e., the elongation rate) is between those values with 95% confidence.

What's with the confidence intervals for $E(y_{new})$ (i.e., stem length) for a given day of the year? That is, what is the confidence interval of $E(y_{new}|\mathbf{x})$?\
We are looking at\
$$\hat{\beta_0}+\hat{\beta_1}x_{new} \pm t_{dfe, 1-\alpha/2}s.e. (\hat{\beta_0}+\hat{\beta_1}x_{new}),$$ where $s.e.(\hat{\beta_0}+\hat{\beta_1}x_{new}) = \hat{\sigma}\sqrt{\frac{1}{n}+ \frac{(x-\hat\mu_x)^2}{(n-1)s_x^2}}$.

**Note**: Remember that $E(y)$ is the **mean** of all potentially observable values of $y$. That is why this interval is somewhat smaller than the prediction interval.

```{r}
doy <- seq(min(data$doy), max(data$doy), by = 1)
estim <- predict(m, data.frame(doy), interval="confidence")

fitted.data <- bind_cols(estim, doy =doy) 

fitted.data %>% 
  ggplot(aes(doy, fit))+
  geom_point(aes(x = doy, y = stm.length_cm), data = data, color = 'grey30')+
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = 'grey60', alpha = .3)+
  geom_line()+
  labs(x = "Day of the year", 
       y = "Stem length (cm)")+
  theme_classic()+
  theme(aspect.ratio = 1)
```

#### Prediction uncertainty

What's with the confidence intervals for $y$ (i.e., stem length) for a given day of the year? What is the confidence interval of $\mathbf{y}|\mathbf{x}$?\
Previously, we only had $\text{Var}(\hat\beta)$. Now we have to include $\text{Var}(\hat\varepsilon)$:

$$s.e.y_{new}=\hat\sigma \sqrt{1+\frac{1}{n}+ \frac{(x-\hat\mu_x)^2}{(n-1)s_x^2}}$$

```{r}
pred <- predict(m, data.frame(doy), interval="prediction")

fitted.data <- bind_cols(fitted.data, 
                         pred_lwr = pred[,2], 
                         pred_upr = pred[,3]) 

fitted.data %>% 
  ggplot(aes(doy, fit))+
  geom_point(aes(x = doy, y = stm.length_cm), data = data, color = 'grey30')+
  geom_ribbon(aes(ymin = pred_lwr, ymax = pred_upr), fill = 'grey85', alpha = .3)+
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = 'grey60', alpha = .3)+
  geom_line()+
  labs(x = "Day of the year", 
       y = "Stem length (cm)")+
  theme_classic()+
  theme(aspect.ratio = 1)
```

## Uncertainty as affected by the amount of data

-   Thought experiment: what happens to both uncertainties when $n$ is infinite? (what happens to $\text{Var}(\hat\beta)$ when $n$ is infinite?)\
-   Whiteboard: analyze $\text{Var}(\hat\beta) = \frac{\sigma^2}{(n-1)s^2_x}$ when $n$ is infinity.

```{r}
set.seed(20)
dd <- read.csv(url)[sample(1:nrow(data), size = 9, replace = FALSE),]
m <- lm(stm.length_cm ~ 1 + doy,  data = dd)
doy <- seq(min(dd$doy), max(dd$doy), by = 1)
estim <- predict(m, data.frame(doy), interval="confidence")

fitted.data <- bind_cols(estim, doy =doy) 
pred <- predict(m, data.frame(doy), interval="prediction")

fitted.data <- bind_cols(fitted.data, 
                         pred_lwr = pred[,2], 
                         pred_upr = pred[,3]) 

fitted.data %>% 
  ggplot(aes(doy, fit))+
  geom_point(aes(x = doy, y = stm.length_cm), data = dd, color = 'grey30')+
  geom_ribbon(aes(ymin = pred_lwr, ymax = pred_upr), fill = 'grey85', alpha = .3)+
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = 'grey60', alpha = .3)+
  geom_line()+
  labs(x = "Day of the year", 
       y = "Stem length (cm)")+
  theme_classic()+
  theme(aspect.ratio = 1)
```

[In-class R code](https://github.com/jlacasa/stat705_fall2024/tree/main/classes/in_class_code/day7_uncertainty.qmd)

## For next class/later today

-   Resubmit your assignments today!\
-   Questions about the project? Proposals are due next Friday (09/13)!
