---
title: "Day 8 - 09/06/2024"
format: html
editor: visual
---

## From last class

-   Be suspicious of point estimates that are reported alone.  

## Uncertainty  

### Intro - data/model prep  

```{r message=FALSE, warning=FALSE}
library(tidyverse)
url <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/corn_example2.csv"
data <- read.csv(url)
data %>% 
  ggplot(aes(plant_density, yield_Mgha))+
  geom_point()+
  labs(x = expression(Plant~Density~(plants~m^{-2})), 
       y = expression(Grain~Yield~(Mg~ha^{-1})))+
  theme_classic()+
  coord_cartesian(xlim = c(0, 15), 
                  ylim = c(0, 16), 
                  expand = F)+
  theme(aspect.ratio = 1)
```

What would be a good statistical model for these data?  

The yield from the $i$th observation, $y_i$, can be described as 
$$y_i \sim N(\mu_i , \sigma^2),$$

$$\mu_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 ,$$
for $i =1, 2, ..., n$, where $n$ is the total number of observations, $\mu_i$ is the mean of the $i$th observation, $\beta_0$ is the intercept (i.e., the yield with zero plants), $\beta_1$ and $\beta_2$ determine the shape of the quadratic curve, $x_i$ is the plant density of the $i$th observation, and $\sigma^2$ is the variance of the data.  

```{r}
m <- lm(yield_Mgha ~ plant_density + I(plant_density^2), data = data)
summary(m)
```


### Estimation uncertainty  


```{r}
data_new <- data.frame(plant_density = 10)
```

Let's look at 
$$E(y_i) = E(\beta_0 + \beta_1 x_i + \beta_2 x_i^2).$$
```{r}
predict(m, data_new, type = "response")
```


Let's look at the 95% CI for 
$$E(y_i) = E(\beta_0 + \beta_1 x_i + \beta_2 x_i^2).$$

```{r}
predict(m, data_new, interval = "confidence")
```

### Prediction uncertainty  

Let's look at the 95% CI for 
$$y_i = \beta_0 + \beta_1 x_i + \beta_2 x_i^2 + \varepsilon_i,$$
where $\varepsilon_i = y_i- \mu_i$.


```{r}
predict(m, data_new, interval = "prediction")
```

### Uncertainty of derived quantities  

- What is a derived quantity?  
- Invariance property of Maximum likelihood estimators.  
- What options are there out there to calculate the uncertainty of a derived quantity?  
- Check out [this paper](https://link.springer.com/article/10.1186/s13007-024-01261-9).  

#### The 'optimum planting rate'  

The first derivative of $\beta_0 + \beta_1 x_i + \beta_2 x_i^2$ is $\beta_1 + 2x_i \beta_2$. If we set that to zero, 

$$\beta_1 + 2x_i \beta_2 = 0, $$

and 

$$OPR = x_i  = - \frac{\beta_1}{2 \beta_2}.$$

Using the invariance property of MLE, we can obtain the MLE for this derived quantity.  

```{r}
coef(m)
(opd_hat <- -coef(m)[2]/(2*coef(m)[3]))
```

What would be a reasonable confidence interval for this derived quantity?   

#### Can we apply the invariance propert and use the 95% CI from the $\beta$s to estimate the 95CI for the derived quantity?  


```{r}
confint(m)
```

```{r}
covariance <- vcov(m)
covariance
```

[Remember the simulation?](https://jlacasa.github.io/stat705_fall2024/classes/in_class_code/day8_simulation_example) [[R code](https://github.com/jlacasa/stat705_fall2024/tree/main/classes/in_class_code/day8_simulation_example.qmd)]


#### The delta method  


```{r}
opd_se_hat <- msm::deltamethod(g = ~ -x2/(2*x3), mean = coef(m), cov = covariance)
opd_hat - 1.96*opd_se_hat ## 1.96 is typically used for approximate confidence intervals
opd_hat + 1.96*opd_se_hat
```

## Summary    

- Why are uncertainty estimates important?  
- What are uncertainty estimates telling us?  
- Which variables have uncertainty? Which ones don't?  
- Where does uncertainty come from?  

## For next week  

- [Assignment 2](https://jlacasa.github.io/stat705_fall2024/assignments/hw2) is up.  
- Semester project proposals are due next Friday.
- [An example proposal](https://github.com/jlacasa/stat705_fall2024/blob/main/homeworks/finalproj.pdf). 

