---
title: "Day 5 - 08/28/2024"
format: html
editor: visual
---

## From last class

-   Please submit Assignment 1 today!

-   **A few things to review**

    -   What is an unbiased estimator?\
        When we use $\hat{\beta}$ as an estimator of $\beta$, there are no systematic mechanisms in the computation of $\hat{\beta}$ that will lead to over/underestimation of $\beta$. That is why we expect to obtain $\beta$ (i.e., the "true" $\beta$) back.

## Estimation

### Point estimates in $\boldsymbol{\beta}$

```{r message=FALSE, warning=FALSE}
library(tidyverse)
dd_lotus <- read.csv("data/lotus_part1.csv") %>% 
  transmute(species = factor(species), agb_g) 
summary(dd_lotus)
m1 <- lm(agb_g ~ 0 + species, data = dd_lotus)
```

Let's calculate $\hat{\boldsymbol{\beta}}_{LSE} = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}$

```{r}
X <- model.matrix(m1)
y <- dd_lotus$agb_g
(beta_hat <- solve(t(X)%*%X) %*% t(X)%*%y)
```

```{r}
coefficients(m1)
```

### Residuals

```{r}
residuals <- y - X%*%beta_hat
```

### Variance $\sigma^2$

-   Perhaps one of the most important parameters we will estimate.\
-   Prediction accuracy.
-   Prediction intervals, estimation intervals.
-   Degrees of freedom. From $n-1$ to $n-p$.

Let's calculate $\hat{\sigma}^2 = \frac{\boldsymbol{\hat{\varepsilon}^T\hat{\varepsilon}}}{n-p} = \frac{RSS}{n-p}$

```{r}
RSS <- sum(residuals^2)
n <- nrow(dd_lotus)
p <- length(beta_hat)
(sigma2_hat <- RSS/(n-p)) 
(sigma_hat <- sqrt(sigma2_hat))
```

```{r}
summary(m1)$sigma
```

### Confidence intervals

Confidence intervals are a measure of uncertainty. They are related to other measures of uncertainty, like $\sigma^2$. We know that $$CI = \hat{\beta} \pm t_{dfe, 1 - \alpha/2} \ s.e.,$$ where $t_{\alpha/2, dfe}$ is the t-score that corresponds to the confidence level (i.e., $\alpha$) with the degrees of freedom of the error $\text{rdf}$, and $s.e.$ is the standard error.

-   Why t distribution?

-   Let's calculate the standard error:

```{r}
unscaled_covariance <- solve(t(X)%*%X)
```

```{r}
(se <- sqrt(diag(unscaled_covariance))*sigma_hat)
```

```{r}
alpha <- 0.05
dfe <- n-p
(t_score <- qt(p=1-alpha/2, df = dfe, lower.tail=T))
```

```{r}
se*t_score
```

```{r}
data.frame(lower = beta_hat - se*t_score, 
           upper = beta_hat + se*t_score)
```

```{r}
confint(m1)
```

#### Interpretation

-   "We are 95% confident that ..."

### In `summary`:

```{r}
summary(m1)
```

### Let's do the same, from scratch, for continuous data

Live R code [here](https://github.com/jlacasa/stat705_fall2024/blob/main/classes/in_class_code/day5_estimation_example.qmd).

### How much money would you bet on the estimate? Let's do a simulation.

R code [here](https://github.com/jlacasa/stat705_fall2024/blob/main/classes/in_class_code/day5_simulation_example.qmd).

## For next class

-   No reading!
-   Assignment 1.
-   Take a look at the schedule.
