---
title: "Day 6 - 08/30/2024"
format: html
editor: visual
---

## From last class

-   **Assignments**. A few comments:
    -   Good overall.\
    -   `geom_smooth()` is forbidden from now on! Use `stat_function()` instead.\
    -   Resubmit next Wednesday.
    -   How is a statistical model built.
        -   Deterministic function
        -   Likelihood function
        -   Method of estimation

## Linear models revisited

Let's look at the clover data again.

```{r}
url <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/lotus_hw1.csv"
data <- read.csv(url)
```

```{r include=FALSE, eval=TRUE}
library(tidyverse)
```

```{r}
data %>% 
  ggplot(aes(doy, stm.length_cm))+
  geom_point()+
  labs(x = "Day of the year", 
       y = "Stem length (cm)")+
  theme_classic()+
  theme(aspect.ratio = 1)
```

What's a good model for the data?

One option is $$y_i = \beta_0 + x_i \beta_1 + \varepsilon_i, $$ $$\varepsilon_i \sim N(0, \sigma^2),$$

for $i = 1, 2, ..., n$ (n being the total number of observations), where $y_i$ is the length of the stem (in cm) of the $i$th observation, $\beta_0$ is the length of the stem (in cm) at day of the year 0 (i.e., December 31 of last year), $x_i$ is the day of the year of the $i$th observation, $\varepsilon_i$ is the residual of the $i$th observation, that is normally distributed.

This is the same as writing

$$\mathbf{y} \sim \text{N}(\boldsymbol{\mu}, \sigma^2\mathbf{I}),$$ $$\boldsymbol{\mu}=\mathbf{X}\boldsymbol{\beta}.$$

### Estimation

Let's fit the model written above to data. [R script](https://github.com/jlacasa/stat705_fall2024/blob/main/classes/in_class_code/day6_estimation_example.Rmd)

```{r include=FALSE, eval=TRUE}
m_cont <- lm(stm.length_cm ~ 1 + doy,  data = data)
```

#### Confidence intervals

```{r}
confint(m_cont)
```

**Some interpretations**

-   The difference in stem length for two consecutive days is between `r round(confint(m_cont)[2], 2)` and `r round(confint(m_cont)[4], 2)` with 95% confidence. An additional day growing is associated with and increase in **average** stem length that is between `r round(confint(m_cont)[2], 2)` and `r round(confint(m_cont)[4], 2)` with 95% confidence.\[[Source (Chapter 3, page 78)](https://www.routledge.com/Understanding-Regression-Analysis-A-Conditional-Distribution-Approach/Westfall-Arias/p/book/9780367493516?srsltid=AfmBOoqnIcDRh-a5BdDe-WJfoKZRLmZd5sCEf8P52su9Jm3rLo28gRO3)\].

-   Also: the interval (`r round(confint(m_cont)[2], 2)`, `r round(confint(m_cont)[4], 2)`) contains all the values $\beta_1^{\star}$ where we would fail to reject the null hypothesis that $\beta_1 = \beta_1^{\star}$ at level $\alpha$. \[[Source](https://doi.org/10.1890/13-0590.1)\].

### How much money would you bet on the estimate? Let's do a simulation.

R code [here](https://github.com/jlacasa/stat705_fall2024/blob/main/classes/in_class_code/day5_simulation_example.qmd).

## For next class

-   No class on Monday.\
-   Review your assignments.
