---
title: "Day 15 - 09/23/2024"
format: html
editor: visual
---

## Analysis of Variance: ANOVA   

Helps us answering objectives like finding out "is A different to B or C?", or "do any of these treatments have an effect on the mean?  

Another types of objective (that require other types of modeling tools), for example, would be "to predict y for given levels of A and B" (e.g., predict corn yield with 150 kg N per ha and 20 inches rainfall in the season).

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggpubr)
url <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/lotus_part4.csv"
dd <- read.csv(url)

dd_subset <- dd %>% 
  filter(trt == "ctrl")
```


```{r message=FALSE, warning=FALSE}
dd %>% 
  ggplot(aes(species, crown_g))+
  geom_boxplot(aes(fill = species))+
  scale_fill_manual(values = c("#DB504A", "#43AA8B", "#FAD4C0"))+
  labs(x = "Species", 
       y = "Crown biomass (g)", 
       fill = "Species")+
  theme_classic()+
  theme(aspect.ratio = 1)
```

Let's fit the model $$y_{ij} \sim N(\mu_{ij}, \sigma^2),$$

$$\mu_{ij} = \mu + \beta_{0j},$$ where $y_{ij}$ is the observation of crown biomass for the $i$th observation of the $j$th species, $\mu{ij}$ is its mean and $\sigma^2$ its variance. The means is described with the overall mean $\mu$ and $\beta_{0j}$, the effect of the $j$th species.

```{r}
m1 <- lm(crown_g ~ species, data = dd_subset)
summary(m1)
```

What's another way of writing out this statistical model?  

[Live R coding.](https://github.com/jlacasa/stat705_fall2024/blob/main/classes/in_class_code/day15_inclass.Rmd) 


```{r}
model.matrix(m1)[1:30,]
```

### One-way ANOVA   

$$H_0: y_{ij} = \mu + \varepsilon_{ij}$$
($H_0:$ All means are the same regardless of the predictor variable, indicated in subscript $j$).

$$H_1: y_{ij} = \mu + \beta_j + \varepsilon_{ij}$$
($H_1:$ The means are different depending on the predictor variable, indicated in subscript $j$).

```{r}
car::Anova(m1, type = "II")
car::Anova(m1, type = "III")
```

::: {.alert .alert-info}
-   <strong>Self-study questions:</strong> 
  - If the treatment factor has two levels, what is the difference between a t-test and an ANOVA test?   
  - Can you know which mean is larger based on the ANOVA results?  
  
:::

### Two-way ANOVA  

Used to answer the same types of questions, this analysis is most common in factorial designs.  


```{r}
dd %>% 
  ggplot(aes(species, crown_g))+
  geom_boxplot(aes(fill = trt, color = species))+
  scale_fill_manual(values = c("white", "#F7B05B"))+
  scale_color_manual(values = c("#DB504A", "#43AA8B", "#FAD4C0"))+
  labs(x = "Species", 
       y = "Crown biomass (g)", 
       fill = "Treatment", 
       color = "Species")+
  theme_classic()+
  theme(aspect.ratio = 1)
```

```{r}
m2 <- lm(crown_g ~ species*trt, data = dd)
summary(m2)
```

```{r}
model.matrix(m2)[1:30,]
```

```{r}
car::Anova(m2, type = "II")
car::Anova(m2, type = "III")
```


## Using a combination of categorical (indicator) variables and continuous variables  

### Review types of explanatory data    

```{r}
url2 <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/lotus_part3.csv"
dd2 <- read.csv(url2) 
ggarrange(
 dd2 %>% 
   filter(doy ==237) %>% 
    ggplot(aes(species, crown_g))+
    geom_boxplot(aes(fill = species))+
    scale_fill_manual(values = c("#DB504A", "#43AA8B", "#FAD4C0"))+
    labs(x = "Species", 
         y = "Crown biomass (g)", 
         fill = "Species")+
    theme_classic()+
    theme(aspect.ratio = 1), 
 dd2 %>% 
   filter(species== "A") %>% 
   ggplot(aes(doy, crown_g))+
    geom_point(aes(fill = species), shape = 21, size =3.5)+
    scale_fill_manual(values = c("#DB504A", "#43AA8B", "#FAD4C0"))+
    labs(x = "Day of the Year", 
         y = "Crown biomass (g)", 
         fill = "Species")+
    theme_classic()+
    theme(aspect.ratio = 1),
 ncol =2, 
 labels = c("Categorical/Indicators", "Continuous"), 
 label.x = c(0, 0.2)
)
```

```{r}
dd2 %>% 
  ggplot(aes(doy, crown_g))+
  geom_point(aes(fill = species), shape = 21, size =3.5)+
  scale_fill_manual(values = c("#DB504A", "#43AA8B", "#FAD4C0"))+
  labs(x = "Day of the Year", 
       y = "Crown biomass (g)", 
       fill = "Species")+
  theme_classic()+
  theme(aspect.ratio = 1)
```

```{r}
m3 <- lm(crown_g ~ species*doy, data = dd2)

summary(m3)

car::Anova(m3, type = "III")
```

```{r}
model.matrix(m3)[1:30,]
```


## Let's put this in context  

- Don't forget our assumptions:  
  - residuals are iid $\sim N(0, \sigma^2)$  
  - parameters enter linearly (linear model)  
  - homoscedasticity (i.e., constant variance)  
  
- Whiteboard. Context in data analysis workflow.

## For next class  

- Read Chapter 15.  
- Resubmit Assignment 2.
- Following: model diagnostics, model selection. 