---
title: "Day 14 - 09/20/2024"
format: html
editor: visual
---

## From last class

-   Assignment expectations.
    -   Model homework.
    -   Uncertainty estimates.
-   [Resources on the statistical models we're fitting](https://jlacasa.github.io/stat705_fall2024/classes/complementary_model_notation)\
-   Rmd resources:
    -   Rmd cookbook \[[link](https://bookdown.org/yihui/rmarkdown-cookbook/)\]
    -   Rmd cheatsheet \[[link](https://rstudio.github.io/cheatsheets/html/rmarkdown.html?_gl=1*ianen8*_ga*OTc0NDYwNTUxLjE3MjQ2MDI3Nzk.*_ga_2C0WZ1JHG0*MTcyNjc4MjQyNS43LjEuMTcyNjc4MjQ4Ni4wLjAuMA..)\]\
    -   R (tidyverse) YouTube guide \[[link](https://www.youtube.com/playlist?list=PLLxj8fULvXwGOf8uHlL4Tr62oXSB5k_in)\]

## Effects, interactions, notation

```{r message=FALSE, warning=FALSE}
library(tidyverse)
url <- "https://raw.githubusercontent.com/jlacasa/stat705_fall2024/main/classes/data/lotus_part3.csv"
dd <- read.csv(url)

dd %>% 
  ggplot(aes(doy, crown_g))+
  geom_point(aes(fill = species), shape = 21, size =3.5)+
  scale_fill_manual(values = c("#DB504A", "#43AA8B", "#FAD4C0"))+
  labs(x = "Day of the Year", 
       y = "Crown biomass (g)", 
       fill = "Species")+
  theme_classic()+
  theme(aspect.ratio = 1)
```

Let's fit the model $$y_{ij} \sim N(\mu_{ij}, \sigma^2),$$

$$\mu_{ij} = \beta_{0j} + \beta_{1j}x_{ij},$$ where $y_{ij}$ is the observation of crown biomass for the $i$th observation of the $j$th species, $\mu{ij}$ is its mean and $\sigma^2$ its variance. The means is described with $\beta_{0j}$, the intercept of the $j$th species and $\beta_{1j}$, the slope of the $j$th species

```{r}
m <- lm(crown_g ~ species * doy, data = dd)
summary(m)
```

It's getting much harder to compute the E(y)s by hand!

```{r}
beta_hat <- coef(m)
beta_hat
```

What's $E(y|x=200, \text{Species A})$?

Compute $\beta_{0A} + 200 \cdot \beta_{1A}$ (or $\beta_{01} + 200 \cdot \beta_{11}$):

```{r}
beta_hat[1] + 200*beta_hat[4] 
```

What's $E(y|x=200, \text{Species C})$?\
Compute $\beta_{0C} + 200 \cdot \beta_{1C}$ (or $\beta_{02} + 200 \cdot \beta_{12}$):

```{r}
(beta_hat[1] + beta_hat[2]) + 200*(beta_hat[4] + beta_hat[5]) 
```

What's $E(y|x=200, \text{Species D})$?\
Compute $\beta_{0D} + 200 \cdot \beta_{1D}$ (or $\beta_{02} + 200 \cdot \beta_{12}$):

```{r}
(beta_hat[1] + beta_hat[3]) + 200*(beta_hat[4] + beta_hat[6]) 
```

Can we use the predict function?

```{r}
X_new <- data.frame(species = c("A", "C", "D"),
                    doy = 200)
predict(m, X_new, interval = "confidence")
```

## Hypothesis tests

In the summary, we're testing if whatever coefficient $\beta = 0$. That means, the null Hypothesis $H_0 : \beta_0=0$, and the alternative Hypothesis $H_1 : \beta_0\ne0$.

```{r}
summary(m)
model.matrix(m)
```

We also tested $H_0: \mu_1 = \mu_2$, $H_1: \mu_1 \ne \mu_2$ using t-tests. What happens when we have more than 2 categories?

## ANOVA

In ANOVA:\
- $H_0: \mu_1=\mu_2=\mu_3$ (all $\mu$s are the same)\
- $H_1$: at least one $\mu$ is different.

### `car::Anova` versus `anova` in R

```{r message=FALSE, warning=FALSE}
library(car)
anova(m)
car::Anova(m, type = 3)
```

[In class code](https://github.com/jlacasa/stat705_fall2024/blob/main/classes/in_class_code/class14.Rmd).

## For next week

-   Read chapter 14 from the book (ANOVA)\
