---
title: "Day 29 - 10/28/2024"
format: html
editor: visual
---

## Mixed Models

- A combination of "fixed effects" and "random effects".  
- Slap a new set of assumptions on top of the ones we had. 
- We estimate fixed effects, $\mu, \boldsymbol{\beta}$, and random effects, $\sigma^2_u, \sigma^2_v, \sigma^2_{\varepsilon}$, etc. 
- Review the assumptions of the general linear model vs. the general linear mixed model.  
- Discuss the type of data needed for a variable with random effects.


### Fixed effects, random effects  

- When should we consider random/fixed effects? There are many opinions about this - see [Gelman (2005), page 20](https://doi.org/10.1214/009053604000001048).   
- A few differences:  
  - Methods of estimation. LSE, shrinkage.   
  - What process is being studied? (interest at the population-level versus at the individual-level)  
  - How many levels does the factor have, vs. how many did we record?  
  - Different types of inference we obtain.  

- [Should blocks be fixed or random?](https://newprairiepress.org/agstatconference/2016/proceedings/4/)  

### Restricted maximum likelihood (REML)  

- REML is the default in most cases because, for small data (aka most cases), maximum likelihood (ML) provides variance estimates that are downward biased.   
- Why is the unbiased estimation of variance components so important? 
  - Relationship between variance estimates, standard error, confidence intervals, t-tests, type I error.  
- In REML, the likelihood is maximized after accounting for the model’s fixed effects.  

- In ML, $-\ell_{ML}(\boldsymbol{\sigma; \boldsymbol{\beta}, \mathbf{y}}) = - (\frac{n}{2}) \log(2\pi)-(\frac{1}{2}) \log (|\mathbf{V}(\boldsymbol\sigma)|) - (\frac{1}{2}) (\mathbf{y}-\mathbf{X}\boldsymbol{\beta})^T[\mathbf{V}(\boldsymbol\sigma)]^{-1}(\mathbf{y}-\mathbf{X}\boldsymbol{\beta})$  
- In REML, $\ell_{REML}(\boldsymbol{\sigma};\mathbf{y}) = - (\frac{n-p}{2}) \log (2\pi) - (\frac{1}{2}) \log (|\mathbf{V}(\boldsymbol\sigma)|) - (\frac{1}{2})log \left( |\mathbf{X}^T[\mathbf{V}(\boldsymbol\sigma)]^{-1}\mathbf{X}|\right) - (\frac{1}{2})\mathbf{r}[\mathbf{V}(\boldsymbol\sigma)]^{-1}\mathbf{r}$, where $p = rank(\mathbf{X})$ and $\mathbf{r} = \mathbf{y}-\mathbf{X}\hat{\boldsymbol{\beta}}_{ML}$.  
  - Start with initial values for $\boldsymbol\sigma$, $\tilde{\boldsymbol{\sigma}}$.  
  - Compute $\mathbf{G}(\tilde{\boldsymbol{\sigma}})$ and $\mathbf{R}(\tilde{\boldsymbol{\sigma}})$.  
  - Obtain $\boldsymbol{\beta}$ and $\mathbf{b}$.   
  - Update $\tilde{\boldsymbol{\sigma}}$.  
  - Repeat until convergence.  

### An applied example  

- Review [last class](https://jlacasa.github.io/stat705_fall2024/classes/day28_10252024) (shrinkage)  
- Randomized complete block designs  

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(agridat)
```

```{r}
data(gilmour.serpentine)
dd <- gilmour.serpentine
dd <- dd %>% 
  mutate(rowf=factor(row), colf=factor(10*(col-8)))
str(dd)
```

```{r}
polygon <- data.frame(rep = rep(c("R1", "R2", "R3"), each = 4), 
           col = rep(c(.5, 5.5, .5, 10.5, 10.5, 15.5), each =2), 
           row = rep(c(.5, 22.5, 22.5, .5), 3))
dd %>% 
  ggplot(aes(col, row))+
  geom_raster(aes(fill = yield))+
  coord_equal()+
  geom_polygon(aes(group = rep),
               data = polygon,
               fill = NA, color = 'black')+
  rcartocolor::scale_fill_carto_c(palette = "Earth",
                                  name = expression(Yield~(g~m^{-2}))) +
  theme_minimal()+
  labs(x = "Col", 
       y = "Row")
```

```{r message=FALSE, warning=FALSE}
library(nlme)

m_fixed <- lm(yield ~ 0 + gen + rep, data = dd)
m_mixed <- lme(yield ~ 0 + gen,
               random = ~ 1 | rep,
               data = dd)
m_mixed
```

### Modeling assumptions   

```{r}
dd$residuals <- residuals(m_mixed)
hist(dd$residuals)
dd %>% 
  ggplot(aes(col, row))+
  geom_raster(aes(fill = residuals))+
  coord_equal()+
  geom_polygon(aes(group = rep),
               data = polygon,
               fill = NA, color = 'black')+
  rcartocolor::scale_fill_carto_c(palette = "ArmyRose",
                                  direction = -1,
                                  name = expression(Yield~(g~m^{-2}))) +
  theme_minimal()+
  labs(x = "Col", 
       y = "Row")
```

```{r}
library(mgcv)
m_fixed_spatial <- gam(yield ~ 0 + gen + rep + s(col, row, bs = "tp", k =7), 
                       data = dd)
```


```{r}
dd$residuals_gam <- residuals(m_fixed_spatial)
hist(dd$residuals_gam)
dd %>% 
  ggplot(aes(col, row))+
  geom_raster(aes(fill = residuals_gam))+
  coord_equal()+
  geom_polygon(aes(group = rep),
               data = polygon,
               fill = NA, color = 'black')+
  rcartocolor::scale_fill_carto_c(palette = "ArmyRose",
                                  direction = -1,
                                  name = expression(Yield~(g~m^{-2}))) +
  theme_minimal()+
  labs(x = "Col", 
       y = "Row")
```

```{r}
## compare standard errors
summary(m_fixed)$coefficients[1:10,2]
summary(m_fixed_spatial)$se[1:10]
```

