---
title: "Day 30 - 10/30/2024"
format: html
editor: visual
---

## Announcements  

- Rendering quarto htmls  
```{r eval = FALSE}
title: "title_to_assignment"
format: 
  html:
    embed-resources: true
editor: visual
```
- Assignment is due tomorrow 11:59 pm  
- Projects are due November 22 (3.5 weeks)  
  - Send to one of your classmates for peer review
  - Peer review feedback due December 2nd.  
  - Presentations (15 min) the week of finals.  

## Mixed models  

- Review assumptions on the whiteboard.  
- Differences between fixed versus random effects
  - Methods of estimation. LSE, shrinkage.   
  - What process is being studied? (interest at the population-level versus at the individual-level)  
  - How many levels does the factor have, vs. how many did we record?  
  - Different types of inference we obtain.  
- Criteria for fixed versus random effects.  

### Applied examples  

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

#### Random intercept and slope  

```{r message=FALSE, warning=FALSE}
dd <- read.csv("data/student_data.csv") %>% 
  rownames_to_column("student_ID") %>% 
  pivot_longer(cols = c(G1, G2), values_to = "prev_grade", names_to = "term")
```

```{r}
dd %>% 
  ggplot(aes(prev_grade, G3))+
  geom_point()+
  facet_wrap(~school)
```

```{r message=FALSE, warning=FALSE}
library(nlme)
```

```{r}
m1 <- lm(G3 ~ school + prev_grade : school, data = dd)
m2 <- lm(G3 ~ school+ prev_grade : school + student_ID , data = dd)
m3 <- lm(G3 ~ school + prev_grade : school : student_ID , data = dd)
m4 <- lme(G3 ~ prev_grade : school, random = ~ 1|student_ID, data = dd)
m5 <- lme(G3 ~ school + prev_grade : school, random = ~ 0+prev_grade|student_ID, data = dd)
```

```{r}
dd$p1 <- predict(m1, dd)
dd$p2 <- predict(m2, dd)
dd$p3 <- predict(m3, dd)
dd$p4 <- predict(m4, dd)
dd$p5 <- predict(m5, dd)

dd %>%
  ggplot(aes())+
  facet_wrap(~school)+
  geom_point(aes(x =prev_grade, y=G3))+
  geom_point(aes(x=prev_grade, y=p2, group = school), color = 'lightblue')+
  geom_point(aes(x=prev_grade, y=p3, group = school), color = 'orange')+
  geom_point(aes(x=prev_grade, y=p4, group = school), color = 'gold')+
  geom_point(aes(x=prev_grade, y=p5, group = school), color = 'lightgreen')+
  geom_point(aes(x=prev_grade, y=p1), color = 'tomato')
```

#### Designed experiments  

```{r message=FALSE, warning=FALSE}
library(agridat)
```

```{r}
data(gilmour.serpentine)
dd <- gilmour.serpentine
dd <- dd %>% 
  mutate(rowf=factor(row), colf=factor(10*(col-8)))
str(dd)
```

```{r}
polygon <- data.frame(rep = rep(c("R1", "R2", "R3"), each = 4), 
           col = rep(c(.5, 5.5, .5, 10.5, 10.5, 15.5), each =2), 
           row = rep(c(.5, 22.5, 22.5, .5), 3))
dd %>% 
  ggplot(aes(col, row))+
  geom_raster(aes(fill = yield))+
  coord_equal()+
  geom_polygon(aes(group = rep),
               data = polygon,
               fill = NA, color = 'black')+
  rcartocolor::scale_fill_carto_c(palette = "Earth",
                                  name = expression(Yield~(g~m^{-2}))) +
  theme_minimal()+
  labs(x = "Col", 
       y = "Row")
```

```{r message=FALSE, warning=FALSE}
library(nlme)

m_fixed <- lm(yield ~ 0 + gen + rep, data = dd)
m_mixed <- lme(yield ~ 0 + gen,
               random = ~ 1 | rep,
               data = dd)
m_mixed
```

#### Hierarchical models  

- Further interpretations of (nested) random effects.  

## Coming next  

- Assignment is due tomorrow 11:59 pm  
- Projects are due November 22 (3.5 weeks)  
  - Send to one of your classmates for peer review
  - Peer review feedback due December 2nd.  
  - Presentations (15 min) the week of finals.  

