---
title: "Day 33 - 11/06/2024"
format: html
editor: visual
---

## Announcements  

- Assignment review  
- [glmnet documentation](https://glmnet.stanford.edu/)

## Generalized linear model (GLM) 

Necessary steps to define the statistical model:  

- Identify the probability distribution of $y$  
- - Define what your model is focusing on (usually the expected value $E(y)$)  
- State the linear predictor $\eta$  
- Identify a link function that connects $E(y)$ to the linear predictor  

### The general linear model as a special case of the GLM   

- Identify the probability distribution of $y$  
  - $y_i \sim N(\mu_i, \sigma^2)$
- State the linear predictor $\eta$  
  - $\eta_i = \eta + \tau_i$  
- Identify a link function that connects $E(y)$ to the linear predictor  
  - "identity function": $\eta_i = \mu_i$

### The exponential family  

In general, the GLM can model any distribution from the exponential family:  

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggpubr)

n <- 2000
set.seed(42)
snorm <- rnorm(n, mean = 1, sd = sqrt(1))
spois <- rpois(n, lambda = 1)
sbinom <- rbinom(n, size = 12, prob = .3)
sbeta <- rbeta(n, shape1 = 2, shape2 = 4)
st <- rt(n, df = 10, ncp = 1)
sgamma <- rgamma(n, shape = 10, rate = 2)

data.frame(normal = snorm, 
           poisson = spois, 
           binomial = sbinom, 
           beta = sbeta, 
           t = st, 
           gamma = sgamma) %>% 
  pivot_longer(cols = everything()) %>% 
  mutate(name = factor(name, 
                       levels = c("normal", "t", "gamma", "beta",
                                  "poisson", "binomial"))) %>% 
  ggplot(aes(value))+
  geom_histogram()+
  theme_pubclean()+
  facet_wrap(~name, scales = "free_x")+
  theme(aspect.ratio = 1)
```


## Applied examples   

### Modeling counts of insects    

[R script](in_class_code/day33_inclass.Rmd)

```{r message=FALSE, warning=FALSE}
library(agridat)

data(beall.webworms)
dd <- beall.webworms
str(dd)

unique(dd %>% 
         dplyr::select(trt, spray, lead))

dd %>% 
  ggplot(aes(y , trt))+
  geom_jitter(alpha = .25, width = 0.1)+
  geom_boxplot(aes(color = trt, fill = trt), alpha = .4, outliers = FALSE)+
  theme_classic()+
  labs(y = "Treatment",
       x = "Webworm Counts",
       fill = "Treatment", 
       color = "Treatment")+
  scale_fill_manual(values = c("#FC4C02", "#B47978", "#008E97", "#99D17B"))+
  scale_color_manual(values = c("#FC4C02", "#B47978", "#008E97", "#99D17B"))+
  scale_x_continuous(breaks = 0:9)+
  theme(legend.position = "bottom")

```

```{r message=FALSE, warning=FALSE}
library(mgcv)
m1 <- gam(y ~ spray*lead,
          family = poisson(link = "log"),
          data = dd)

# summary(m1)
anova(m1)
```

```{r}

```





