---
title: "Day 34 - 11/08/2024"
format: html
editor: visual
---

## Announcements  

## Generalized linear model (GLM) 

Necessary steps to define the statistical model:  

- Identify the probability distribution of $y$  
- Define what your model is focusing on (usually the expected value $E(y)$)  
- State the linear predictor $\eta$  
- Identify a link function that connects $E(y)$ to the linear predictor $\eta$  

![Page 60 in [Stroup et al. (2024)](https://www.routledge.com/Generalized-Linear-Mixed-Models-Modern-Concepts-Methods-and-Applications/Stroup-Ptukhina-Garai/p/book/9781498755566?srsltid=AfmBOoo-C2dJm7p5SEH4mizPbgCJC_a8nHQ-mvKc91FaplWRSM499OkL)](misc_plots/GLMM_table.jpg)

### Why the model equation form is unfriendly to non-Gaussian distributions   



### The exponential family  

In general, the GLM can model any distribution from the exponential family:  

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggpubr)

n <- 2000
set.seed(42)
snorm <- rnorm(n, mean = 1, sd = sqrt(1))
spois <- rpois(n, lambda = 1)
sbinom <- rbinom(n, size = 12, prob = .3)
sbeta <- rbeta(n, shape1 = 2, shape2 = 4)
st <- rt(n, df = 10, ncp = 1)
sgamma <- rgamma(n, shape = 10, rate = 2)

data.frame(normal = snorm, 
           poisson = spois, 
           binomial = sbinom, 
           beta = sbeta, 
           t = st, 
           gamma = sgamma) %>% 
  pivot_longer(cols = everything()) %>% 
  mutate(name = factor(name, 
                       levels = c("normal", "t", "gamma", "beta",
                                  "poisson", "binomial"))) %>% 
  ggplot(aes(value))+
  geom_histogram()+
  theme_pubclean()+
  facet_wrap(~name, scales = "free_x")+
  theme(aspect.ratio = 1)
```


## Applied examples   

### Modeling counts of insects    

[R script](in_class_code/day34_inclass.Rmd)

```{r message=FALSE, warning=FALSE}
library(agridat)

data(crowder.seeds)
dd <- crowder.seeds %>% 
  unite("trt", c(gen, extract), remove = FALSE)
str(dd)

dd %>% 
  ggplot(aes(trt, germ, group = plate))+
  geom_linerange(aes(xmin=trt, xmax =trt, ymin = germ, ymax = n, group = plate), alpha = .85, 
               position = position_dodge(width = .25))+
  geom_point(aes(fill = germ/n), shape =21, size =2.5, position = position_dodge(width = .25), alpha = .85)+
  theme_classic()+
  scale_fill_viridis_c()+
  labs(y = "Germination (y, n)",
       x = "Genotype", 
       fill = "Germination (%)")+
  theme(legend.position = "bottom")
```

```{r message=FALSE, warning=FALSE}
library(mgcv)
m1 <- gam(cbind(germ, n-germ) ~ extract*gen + s(plate, bs = "re"),
          family = binomial(link = "logit"),
          data = dd)

summary(m1)
```



## Next week  

- Monday: Robert will teach R documentation  
- Monday+Wednesday: work on projects  
- In exactly 2 weeks: deadline to send your project to your classmates
