---
title: "Day 41 - 12/06/2024"
format: html
editor: visual
---

## An applied example based on the questions I got   

### Review linear regression  

- We're describing changes in y with one continuous predictor.  
- Brainstorm examples.   
- Interpretation of slope parameter.  

```{r message=FALSE, warning=FALSE}
library(agridat)
library(tidyverse)

data(hessling.argentina)
dd <- hessling.argentina
m1 <- lm(yield ~ t10, data = dd)
summary(m1)
```

### Review multiple linear regression  

- We're describing changes in y with multiple continuous predictors.  
- Brainstorm examples.   
- [In class example](in_class_code/day41_inclass.Rmd).
- Interpretation of slope parameters.  
- Multiple regression is actually a good alternative to bad choices [[link](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1046/j.1365-2656.2002.00618.x)].


### Model diagnostics review  

A simple linear mode:  

$$y_{ij} \sim N(\mu_{ij}, \sigma^2),$$

$$\mu_{ij} = \mu + \tau_i + b_j,$$ 

where:

- $y_{ij}$ is the observed yield of the ith genotype and jth repetition, 
- $\mu_{ij}$ is the expected value and $\sigma^2$ the variance, 
- $\mu$ is the overall mean, 
- $\tau_i$ is the effect of the ith genotype, 
- $b_j$ is the effect of the jth block.  

- Review assumptions.  
- Alliance wheat trial example ([Stroup et al., 1994](https://acsess.onlinelibrary.wiley.com/doi/abs/10.2135/cropsci1994.0011183x003400010011x)).  

```{r}
data(gilmour.serpentine)
dd <- gilmour.serpentine
dd <- dd %>% 
  mutate(rowf=factor(row), colf=factor(10*(col-8)))
```

```{r}
polygon <- data.frame(rep = rep(c("R1", "R2", "R3"), each = 4), 
           col = rep(c(.5, 5.5, .5, 10.5, 10.5, 15.5), each =2), 
           row = rep(c(.5, 22.5, 22.5, .5), 3))
dd %>% 
  ggplot(aes(col, row))+
  geom_raster(aes(fill = yield))+
  coord_equal()+
  geom_polygon(aes(group = rep),
               data = polygon,
               fill = NA, color = 'black')+
  rcartocolor::scale_fill_carto_c(palette = "Earth",
                                  name = expression(Yield~(g~m^{-2}))) +
  theme_minimal()+
  labs(x = "Col", 
       y = "Row")
```

```{r}
m1 <- lm(yield ~ gen + rep, data = dd)
```

#### Checking for constant variance  

```{r}
plot(m1$fitted.values, abs(m1$residuals))
```

#### Checking for normal residuals

```{r}
car::qqPlot(m1)
```

#### Checking for independent residuals  

```{r}
dd$residuals <- m1$residuals

library(gstat)
library(sp)
library(latex2exp)

dd_sp <- dd
coordinates(dd_sp) <-  ~ col + row # x+y
v <- variogram(residuals ~ 1, data = dd_sp, cloud=F)

plot(v)

v_fit <- fit.variogram(v, vgm("Sph"))

plot(v,v_fit, cutoff = 10, add =TRUE)
```


```{r}
dd %>% 
  ggplot(aes(col, row))+
  geom_raster(aes(fill = residuals))+
  coord_equal()+
  geom_polygon(aes(group = rep),
               data = polygon,
               fill = NA, color = 'black')+
  rcartocolor::scale_fill_carto_c(palette = "Fall",
                                  name = TeX("$\\epsilon_i$ (g $m^{-2}$)")) +
  theme_minimal()+
  labs(x = "Col", 
       y = "Row")
```

